Effectful Semantics for Substructural SSA

### 2023-11-07 pldi24/paper.tex
Tap

\subsection{Basic Models}

@@

\TODO{find better notation than \(\Pi\)?}

\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) is a pair \((\Pi, \cdot)\) where \(\Pi: M^\omega \to I\) is a map from streams of \(M\) to \(I\) and \(\cdot: M \to I \to I\) is a monoid action such that
  \(
    m \cdot \Pi \ell = \Pi (m: \ell) 
  \)
  where \(m:\ell\) denotes the stream formed by prefixing \(m\) to \(\ell\).
\end{definition}
Two important examples include:
\begin{itemize}
  \item For any monoid \(M\), the unique maps \(\Pi: (\nats \to M) \to \mb{1}\), \(\cdot: M \to \mb{1} \to \mb{1}\) form a stream action on \(\mb{1}\)
  \item The concatenation map \(\Pi: (A^*)^\omega \to A^\omega\) and the concatenation action \(\cdot: A^* \to A^{\leq \omega} \to A^{\leq \omega}\) form a stream action of the monoid of lists of \(A\), \(A^*\), on the set of potentially infinite sequences of \(A\), \(A^{\leq \omega}\).
\end{itemize}

@@

\begin{definition}[Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Pi\;A = A \times M + I\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with
  \[f^\dagger(a) = ...\]
\end{definition}

\begin{definition}[Nondeterministic Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\{\iota_0 (a, 1)\}\)
  and \(\mu_A = ...\)
  This is an Elgot monad, with
  \[f^\dagger(a) = ...\]
\end{definition}

@@

\TODO{relevant = deterministic}

@@

@@

\subsection{Separation Logic}

>>>>

\subsection{Trace Models}

@@

\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to I\) is a map from streams of \(M\) to \(I\) satisfying
  \(
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the stream \(\ell\).
\end{definition}
Some basic examples include:
\begin{enumerate}
  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
\end{enumerate}
We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\). For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq \omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) = \{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains the unit of \(M\).

@@

\begin{definition}[Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Pi\;A = A \times M + I\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\).
\end{definition}

@@

\TODO{relevant = idempotent}

@@

\begin{definition}[Nondeterministic Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\{\iota_0 (a, 1)\}\)
  and \(\mu_A = ...\)
  This is an Elgot monad, with
  \[f^\dagger(a) = \{\} \cup \{\} \cup \{\}\]
\end{definition}

\TODO{relevant = deterministic + idempotent; inclusion of trace monad}

\subsection{Heaps and Separation Logic}

\TODO{equality vs \cite{mellies-ftrs}}
\TODO{potential solutions}

@@
